---
title: "chi massage 2020-now vs 2015-2019"
output: html_notebook
---

load libraries:
```{r}
library("tidyverse")
library("lubridate")
library("janitor")
library("forcats")
library("rvest")
library("ggplot2")
library("readr")
```

load datasets:
```{r}
chibiz2 <- read_csv("chibiz1.csv")
print(chibiz2)
chibizold <- read_csv("chibizold.csv")
print(chibizold)
```

data cleaning:
```{r}
chibiz2 <- chibiz2 %>% clean_names() #standardize column names
chibiz2 %>% filter(duplicated(.)) #check duplicate values
chibiz2 <- chibiz2 %>% distinct() #remove duplicate values
chibiz2 <- chibiz2 %>% mutate(date_issued_new = mdy(date_issued)) #change into date column based on date format presented in dataset
chibiz21 <- chibiz2 %>% select(legal_name, doing_business_as_name, ward, business_activity_id, business_activity, license_number, license_code, application_type, date_issued_new, license_term_expiration_date, license_status, latitude, longitude)
chibiz21 %>% filter(is.na(license_status)) #check for null in specific column
chibizfull <- chibiz21 %>% filter(!is.na(business_activity_id)) 
chibizna <- chibiz21 %>% filter(is.na(business_activity_id))
```

Finding massage parlor licenses:
```{r}
chimassage <- chibizfull %>% filter(grepl("745", business_activity_id))
chimassage %>% select(date_issued_new)
chimassage <- chimassage %>% separate(date_issued_new, into = c("issued_year", "issued_month", "issued_day"), sep = "-")
chi11massage <- chimassage %>% filter(ward == "11")
chi11massage %>% filter(duplicated(license_number))
chi11massage <- chi11massage %>% mutate(license_term_expiration_date = mdy(license_term_expiration_date))
chi11massage <- chi11massage %>% separate(license_term_expiration_date, into = c("expire_year", "expire_month", "expire_day"), sep = "-")
chi11massage
```
From 2020 to now, Chicago has issued/renewed 66 licenses for businesses providing full body massage services in 11th ward. 

Licenses for massage businesses by year:
```{r}
chi11massage %>% filter(issued_year == "2020")
chi11massage %>% group_by(issued_year) %>% count(application_type)
massagetrendnew <- chi11massage %>% group_by(issued_year) %>% count(application_type)
```

check from null table:
```{r}
chibizna %>% filter(ward == "11") %>% group_by(license_code) %>% count()
```
No massage businesses in this dataset. 

currently active massage biz in 11th ward:
```{r}
chi11massage$expire_year <- as.numeric(chi11massage$expire_year)
chi11massage$expire_month <- as.numeric(chi11massage$expire_month)
chi11massage %>% filter(expire_year > 2025)
chi11massage26on <- chi11massage %>% filter(expire_year > 2025)
chi11massage26on %>% group_by(expire_year) %>% count(expire_month)
```
At the end of August 2026, there are going to be 18 currently active licenses for businesses providing full body massage services in the 11th ward. 

export for viz:
```{r}
chi11massage26on$expire_date <- as.Date(with(chi11massagenow, paste(expire_year, expire_month, expire_day,sep="-")), "%Y-%m-%d")
chi11massage26on %>% filter(expire_date > "2026-8-31")
chi11massagecurrent <- chi11massage26on %>% filter(expire_date > "2026-8-31")
chi11massagecurrent %>% filter(!duplicated(license_number))
chi11massagecurrent <- chi11massagecurrent %>% filter(!duplicated(license_number))
write_csv(chi11massagecurrent, "currentchi11massage.csv")
```





chibizold cleaning:
```{r}
chibizold <- chibizold %>% clean_names()
chibizold %>% group_by(ward) %>% count()
chibizold %>% filter(duplicated(.)) #check duplicate values
chi11bizold <- chibizold %>% distinct() #remove duplicate values
chi11bizold <- chi11bizold %>% mutate(date_issued_new = mdy(date_issued)) #change into date column based on date format presented in dataset
chi11bizold <- chi11bizold %>% separate(date_issued_new, into = c("issued_year", "issued_month", "issued_day"), sep = "-")
chi11bizold %>% group_by(issued_year) %>% count()
chi11bizoldfull <- chi11bizold %>% filter(!is.na(business_activity_id)) 
chi11bizoldna <- chi11bizold %>% filter(is.na(business_activity_id))
```

finding massage biz in old dataset:
```{r}
chi11bizoldfull %>% filter(grepl("745", business_activity_id)) 
chi11bizoldfull %>% filter(grepl("745", business_activity_id)) %>% group_by(application_type) %>% count(issued_year)
chi11bizoldna %>% group_by(license_code) %>% count()
chi11bizoldna %>% filter(license_code == "4404")
chi11bizold %>% filter(license_number == "2563893" | grepl("745", business_activity_id))
chi11massageold <- chi11bizold %>% filter(license_number == "2563893" | grepl("745", business_activity_id))
```
From 2015 to 2019, Chicago issued/renewed 76 licenses to businesses providing full body massage services in the 11th ward. 

licenses for massage businesses by year in old dataset:
```{r}
chi11massageold %>% group_by(issued_year) %>% count(application_type)
massagetrendold <- chi11massageold %>% group_by(issued_year) %>% count(application_type)
```

export for viz:
```{r}
write_csv(massagetrendold, "massagetrendold.csv")
write_csv(massagetrendnew, "massagetrendnew.csv")
```




